Bayesian regression facilitates quantitative modelling of cell metabolism

### 2023-10-31 paper/main.qmd
paper update

The training testing split was selected to achieve a large difference between
the fluxes of the training and testing dataset. The split was determined as
we are interested in showing how our model can fit to varied conditions, and
conditions closer to the training set are likely to be predicted well without
necessarily learning the system.

@@

The prior distributions and corresponding true parameter values used in our
case study are shown in supporting information section 3.2. They were chosen to
reflect a plausible pre-experimental information state. In 7 cases, the marginal
prior distribution for a parameter disagrees with the true parameter values used
to generate the data. A similar situation is likely to occur in practice due to
*in vivo* vs *in vitro* measurement differences.

>>>>

The split between training and validation conditions was selected to achieve a
large difference in fluxes between the two categories. The split was determined
as we are interested in showing how our model can fit to varied conditions, and
conditions closer to the training set are likely to be predicted well without
necessarily learning the system.

@@

The prior distributions and corresponding true parameter values used in our
case study are shown in supporting information section 3.2. They were chosen to
reflect a plausible pre-experimental information state. In 7 cases, the marginal
prior distribution for a parameter disagrees with the true parameter values used
to generate the data; a similar situation is likely to occur in practice due to
*in vivo* vs *in vitro* measurement differences.

### 2023-10-31 paper/main.tex
paper update

\title[]{Bayesian regression facilitates quantitative modelling of cell
metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Our paper reviews the existing options in this
area, presents a case study illustrating how Maud can be used to analyse
a metabolic network and explains the biological, statistical and
computational design decisions underpinning Maud.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi

@@

Another advantage is that Bayesian inference problems are well-posed
even when not all parameters are strongly identified. Sloppy models in
which measurable quantities are sensitive to combinations of parameters
but not to individual marginal parameter values are ubiquitous in models
of biological systems \citep{gutenkunst_2007, white_2016}. The parameter
correlation structure represents the set of potential models that
describe the observed data. As we demonstrate in our case study below,
capturing this correlation structure is difficult outside a Bayesian
context.

@@

There have also been efforts to implement Bayesian inference for kinetic
models without the use of MCMC. Examples of alternative inference
methods include variational inference as in
\citet{st.johnBayesianInferenceMetabolic2018}, rejection sampling and
approximate Bayesian computation \citet{saa_construction_2016} and
Laplace approximation, in which the Fisher information matrix is used to
calculate a normal approximation around the maximum a posteriori
parameter configuration
\citep{liebermeister_model_2021, gopalakrishnan_k-fit_2020, stapor_pesto_2018, raue_joining_2013}.
Non-MCMC-based Bayesian kinetic models have limited utility because they
lack reliable diagnostic tools for verifying that their results
approximate the target posterior distribution. This is a problem because
realistic kinetic models tend to induce highly correlated, non-Gaussian,
joint probability distributions
\citep{gutenkunst_2007, stapor_pesto_2018}.

Our application Maud is the first Bayesian kinetic model to combine
biologically realistic mechanistic accuracy---including accurate rate
laws, post- translational modification and thermodynamics---with fast,
robust MCMC sampling using adaptive Hamiltonian Monte Carlo. Further,
Maud is a general-purpose application that can be used to fit a wide
range of Bayesian kinetic models.

@@

To demonstrate our application's capabilities, we used Maud to analyse
an artificial dataset based on the human methionine cycle. We first used
Maud to generate simulated training and validation measurements based on
plausible parameter values, then performed posterior sampling. Next, we
used Maud to predict the validation data.

To show that Maud is robust to missing measurements we compared the
results of fitting the full dataset with an intentionally incomplete
dataset. To demonstrate why a full Bayesian approach is preferable to an
approach based on a Laplace approximation of the posterior distribution,
we also fit our methionine cycle model using the latter method and
compared the results with MCMC sampling.

Finally, we investigated our results to find out what our model learned
about the contributions of different regulatory factors to the flux
through GNMT, an important reaction. This analysis illustrates how Maud
can be used to generate actionable insights about metabolism without the
need for further statistical analysis.

@@

The prior distributions and corresponding true parameter values used in
our case study are shown in supporting information section 3.2. They
were chosen to reflect a plausible pre-experimental information state.
In 7 cases, the marginal prior distribution for a parameter disagrees
with the true parameter values used to generate the data. A similar
situation is likely to occur in practice due to \emph{in vivo} vs
\emph{in vitro} measurement differences.

>>>>

\title[]{Bayesian regression facilitates quantitative modelling of cell
metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Our paper reviews the existing options in this
area, presents a case study illustrating how Maud can be used to analyse
a metabolic network and explains the biological, statistical and
computational design decisions underpinning Maud.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, interior hidden]}{\end{tcolorbox}}\fi

@@

Another advantage is that Bayesian inference problems are well-posed
even when not all parameters are strongly identified. Sloppy models in
which measurable quantities are sensitive to combinations of parameters
but not to individual marginal parameter values are ubiquitous in models
of biological systems \citep{gutenkunst_2007, white_2016}. The parameter
correlation structure represents the set of potential models that
describe the observed data. As we demonstrate in
Section~\ref{sec-laplace} below, capturing this correlation structure is
difficult outside a fully Bayesian context.

@@

There have also been efforts to implement Bayesian inference for kinetic
models without the use of MCMC. Examples of alternative inference
methods include variational inference
\citep{st.johnBayesianInferenceMetabolic2018}, rejection sampling and
approximate Bayesian computation \citep{saa_construction_2016} and
Laplace approximation, in which the Fisher information matrix is used to
calculate a normal approximation around the maximum a posteriori
parameter configuration
\citep{liebermeister_model_2021, gopalakrishnan_k-fit_2020, stapor_pesto_2018, raue_joining_2013}.
Non-MCMC-based Bayesian kinetic models have limited utility because they
lack reliable diagnostic tools for verifying that their results
approximate the target posterior distribution. This is a problem because
realistic kinetic models tend to induce highly correlated, non-Gaussian,
joint probability distributions
\citep{gutenkunst_2007, stapor_pesto_2018}.

Our application Maud provides the first Bayesian kinetic model to
combine biologically realistic mechanistic accuracy---including accurate
rate laws, post- translational modification and thermodynamics---with
fast, robust MCMC sampling using adaptive Hamiltonian Monte Carlo.
Further, Maud is a general-purpose application that can be used to fit a
wide range of Bayesian kinetic models.

@@

To demonstrate our application's capabilities, we used Maud to analyse
an artificial dataset based on the human methionine cycle. We generated
this dataset using Maud, by simulating measurements for a set of
training and validation conditions based on plausible parameter values.
Next, we performed posterior inference for the training measurements and
prediction for both training and validation measurements.

We investigated Maud's sensitivity to missing measurements by comparing
the results of fitting a full dataset with an intentionally incomplete
dataset. To demonstrate why a full Bayesian approach is preferable to an
approach based on a Laplace approximation of the posterior distribution,
we compared the results of analysing a representative metabolic network
using both methods.

Finally, we dug into our results to find out what our full dataset
methionine model learned about the contributions of different regulatory
factors to the flux through GNMT, an important reaction. This analysis
illustrates how Maud can be used to generate actionable insights about
metabolism without the need for further statistical analysis.

@@

The prior distributions and corresponding true parameter values used in
our case study are shown in supporting information section 3.2. They
were chosen to reflect a plausible pre-experimental information state.
In 7 cases, the marginal prior distribution for a parameter disagrees
with the true parameter values used to generate the data; a similar
situation is likely to occur in practice due to \emph{in vivo} vs
\emph{in vitro} measurement differences.
